---
output:
  pdf_document: default
  html_document: default
---
# Pipeline for Biomarker Discovery with Microarray Data

------------------------------------------------------------------------

## 1. Pre-Processing

This part of the pipeline deals with ensuring the quality of the data for downstream processing in later steps. Normalization and various analysis tasks are conducted to verify quality and whether the data make sense from a biological perspective and thus free from any obvious and not-so obvious technical perturbations.

### 1.1 Load necessary libraries

```{r}
library(GEOquery)
library(affy)
library(dplyr)
library(ggplot2)
library(car)
library(reshape2)        
library(pheatmap)
library(ggrepel)
```

### 1.2 Do some cleanup of already existing data from prior runs

Since data is kept locally and expanded from compressed archives, starting anew makes this step necessary.

```{r}
source("./DeleteFiles.R")
temp_d<-tempdir()
DeleteFiles(tempdir(), c("\\.CEL$", "\\.CEL.gz$"))
```

### 1.3 Read the Microarray Data

This includes GEO Series Metada as well as the expression data from the Affymetrix files.

```{r}
geods<-"GSE21942"
tar<-paste(geods, "_RAW.tar", sep="")

untar(tar, exdir=temp_d)
gz_f<-list.files(temp_d, pattern="\\CEL.gz$", full.names=TRUE)
cel_fs<-sapply(gz_f, function(f) {
  R.utils::gunzip(f, remove=FALSE)
})
celfiles<-list.files(temp_d, pattern="\\.CEL$")
affy_data<-ReadAffy(filenames=celfiles, celfile.path=temp_d)
```

### 1.4 Normalize and Extract the Expression Data

```{r}
norm_d<-rma(affy_data)
expressions<-exprs(norm_d)
```

### 1.4 Visualize the Data

This will give an impression about the quality of the data after normalization.

```{r}
boxplot(expressions)
expr_melt<-melt(expressions)
ggplot(expr_melt,aes(x=value, color=Var2)) +
  geom_density() +
  theme_minimal()
hist(expressions)
```

### 1.5 Run PCA

Run PCA and plot the first 2 principal components. Also summarize the PCA analysis results.

```{r}
gse<-getGEO(geods)
eset<-gse[[1]]
pca<-prcomp(t(expressions), scale.=TRUE)
pca_df<-as.data.frame(pca$x)
pca_df$Group<-pData(eset)$`disease state:ch1`
ggplot(pca_df, aes(x=PC1, y=PC2, color=Group, label=rownames(pca_df))) +
  geom_point(size=3) +
  geom_text_repel(size=3) +
  theme_minimal() 
summary(pca)
```

### 1.6 Create Scree Plots for Variance and Cumulative Variance from the PCA analysis

Further depict variance as explained by PCA by plotting appropriate graphs.

```{r}
var_explain<-(pca$sdev)^2 / sum((pca$sdev)^2)
cumsum_var<-cumsum(var_explain)
plot(var_explain, type='b', xlab="PC", ylab="Variance Explained", main="Scree Plot")
plot(cumsum_var, type='b', xlab="PC", ylab="Cumulative Variance Explained", main="Scree Plot - Cumulative Variance")
```

### 1.7 Display Samples Correlation with a Heatmap

The correlation should make sense from the biological perspective. That is, the control samples should cluster together and accordingly the test ones as well. Ideally the correlation should exist only for samples from the same group.

```{r}
annotation<-data.frame(DiseaseStatus=pData(eset)$`disease state:ch1`)
rownames(annotation)<-colnames(expressions)
cor_mat<-cor(expressions)
pheatmap(cor_mat, annotation_col = annotation, 
         clustering_distance_cols = "euclidean", clustering_distance_rows = "euclidean", clustering_method="complete",
         color=colorRampPalette(c("red", "green", "blue"))(50), main="Sample correlation heatmap")
```

## 2. Exploratory Data Analysis

The Exploratory Data Analysis takes the data further down the pipeline by performing and assessing if dimensionality reduction is necessary and if it doesn't distort the data. Various dimensionality reduction methods are evaluated against this rule and investigated upon the delivered results by careful examination of the respective plots.

### 2.1 Install/Load necessary libraries

```{r}
library(umap)
library(uwot)
library(ggplot2)
library(dendextend)
library(factoextra)
library(cluster)
```

### 2.2 Perform dimensionality reduction via UMAP

```{r}
expressions_sample_rows<-t(expressions)
expressions_umap<-umap(expressions_sample_rows)
expressions_umap_df<-as.data.frame(expressions_umap)
expressions_umap_df$Group<-pData(eset)$`disease state:ch1`
ggplot(expressions_umap_df, aes(x=V1, y=V2, colour=Group))+geom_point()
```

### 2.2 Perform hierarchical clustering on reduced data

```{r}
dist_expr<-dist(expressions_umap_df[,1:2], method='euclidean')
hclust_expr<-hclust(dist_expr, method="complete")
dend<-as.dendrogram(hclust_expr)
dend<-color_branches(dend, k=3)
```

### 2.3 Perform hierarchical clustering on non-reduced data

```{r}
expressions_sample_rows_df<-t(expressions)
group<-pData(eset)$`disease state:ch1`
dist_expr_non_reduced<-dist(expressions_sample_rows_df)
hclust_expr_non_reduced<-hclust(dist_expr_non_reduced, method="complete")
dend_non_reduced<-as.dendrogram(hclust_expr_non_reduced)      
dend_non_reduced<-color_branches(dend_non_reduced, k=4)
```

### 2.4 Plot dendrograms for both, reduced and non-reduced data

```{r}
par(mar=c(1,1,1,1))
par(mfrow=c(2,1))
dend_non_reduced %>% set("labels", group) %>% set("labels_colors", as.numeric(as.factor(group)), order_value=TRUE) %>% set("labels_cex", 0.5) %>% plot()
dend %>% set("labels", expressions_umap_df$Group) %>% set("labels_colors", as.numeric(as.factor(expressions_umap_df$Group)), order_value=TRUE) %>% set("labels_cex", 0.5) %>% plot()
```

### 2.5 Calculate and Plot Silhouettes for different values of 'k'

```{r}
fviz_silhouette(silhouette(cutree(hclust_expr, 2), dist_expr))
fviz_silhouette(silhouette(cutree(hclust_expr, 3), dist_expr))
fviz_silhouette(silhouette(cutree(hclust_expr, 4), dist_expr))
fviz_silhouette(silhouette(cutree(hclust_expr, 5), dist_expr))

fviz_silhouette(silhouette(cutree(hclust_expr_non_reduced, 2), dist_expr_non_reduced))
fviz_silhouette(silhouette(cutree(hclust_expr_non_reduced, 3), dist_expr_non_reduced))
fviz_silhouette(silhouette(cutree(hclust_expr_non_reduced, 4), dist_expr_non_reduced))
fviz_silhouette(silhouette(cutree(hclust_expr_non_reduced, 5), dist_expr_non_reduced))
```

## 3. Differential Expression Analysis

### 3.1 Run analysis on Microarray Expressions via Limma

```{r}
library(limma)

meta<-pData(eset)
design<-model.matrix(~ 0 + meta$`disease state:ch1`)
levels(as.factor(make.names(meta$`disease state:ch1`)))
colnames(design)<- meta$`disease state:ch1` %>% make.names %>% as.factor %>% levels

fit<-lmFit(expressions, design)
contr_mat<-makeContrasts(healthy_vs_disease=multiple.sclerosis-healthy, levels=design)
fit2<-contrasts.fit(fit, contr_mat)
fit2<-eBayes(fit2)

results<-topTable(fit2, adjust.method="BH", number=Inf)
head(results)

filtered_results<-results[results$adj.P.Val<=0.05 & abs(results$logFC >= 1), ]
nrow(filtered_results)

results$Group<-meta$`disease state:ch1` %>% make.names
```

### 3.2 Visualize Differential Expression in a Volcano Plot

```{r}
ggplot(results, aes(x=logFC, y=-log10(adj.P.Val))) +
  geom_point(alpha=0.6) +
  geom_hline(yintercept=-log10(0.05),color="blue", linetype="dashed") +
  geom_vline(xintercept=c(-1.0,1.0), color="green", linetype="dashed") +
  labs(title="Expressions from DEA", x="Log2FC", y="-log10(adj.p.val)") +
  theme_minimal()
```

### 3.3 Visualize Changes in Expression over Expression Levels Averaged in an MA Plot

```{r}
limma::plotMA(fit2, main="MA Plot", ylim=c(-5,5))
abline(h = c(-1, 1), col = "blue", lty = 2)
```